---
title: "Project2_EDA"
author: "Hayoung Cheon"
date: "2025-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EDA
## Load data

```{r}
bank <- read.csv("C:/Users/oyunm/Downloads/bank+marketing/bank/bank-full.csv")
head(bank)
```
## boxplot
## contigency table
## hitogram

## Split data into training and validate data set
```{r}
set.seed(1234)
index <- sample(1:nrow(bank), replace=F)
train_size <- floor(0.7 * nrow(bank))

train <- bank[index[1:train_size], ]
test <- bank[index[(train_size + 1):nrow(bank)], ]

```

## Summary stat
```{r}
summary(bank)
```

# outlier?
Outlier in previous data
```{r}
library(tidyverse)
max(bank$previous) #275
bank <- bank %>% filter(previous <= 270)

```

# Variable
```{r}
#Numerical Variables
num.var <- c("age", "balance","day", "duration", "campaign", "pdays", "previous")

#Categorical Variables
bi.categ.var <- c("housing", "loan", "default")

multi.categ.var <- c("job", "marital", "education", "contact", "month", "poutcome")

```

## Key questions
1. Is some variable is a factor of y after accounting for other variables status?  
2. Do any of the factors interact with one another?(Confounding var)  
3. How well can these predictors potentially make a classification/ what are the key factors that contributed to a term deposit.  


## EDA: numerical variables 
```{r}
library(ggplot2)

# Loess curve

# encode y into yes=> 1, no => 0
bank$y.num <- ifelse(bank$y=="yes", 1, 0)

table(bank$y)

# Loop over numeric variables to plot logistic regression curves
for (var in num.var) {
  p <- ggplot(bank, aes_string(x = var, y = "y.num")) +
    geom_point(alpha = 0.3) +
    geom_smooth(
      method = "glm",
      method.args = list(family = "binomial"),
      formula = y ~ x,
      size = 1
    ) +
    ggtitle(paste("Logistic Regression Fit:", var, "vs y.num"))
  print(p)
}


#bank$y.num <- as.numeric(bank$y.num)

#for (var in num.var) {
#  p <- ggplot(bank, aes_string(x = var, y = "y.num")) +
#    geom_point(alpha = 0.4) +
#    geom_smooth(method = "loess", size = 1) +
#    ggtitle(paste("LOESS Curve:", var, "vs y.num"))
#  print(p)
#}


#ggplot(bank, aes(x = duration, y = y.num)) +
# geom_jitter(height = 0.05, alpha = 0.3) +  # jitter helps reveal density in binary data
#  geom_smooth(method = "loess", formula = y ~ x, span = 2.0, size = 1) +
#  ylim(-0.2, 1.2) +
#  labs(title = "LOESS Curve (Binary y): duration vs y.num")



```
**INTERPRETATION**
Increase  
- balance
- duration
- pdays
- previous
- age

Decrease
- campaign

## EDA: categorical variable: binary
```{r}
library(ISLR)
library(epitools)
library(dplyr)
library(tidyverse)

# contigency table
# Ensure y is a factor
bank$y <- as.factor(bank$y)

# Loop over each binary categorical variable to print contingency tables
for (var in bi.categ.var) {
  cat("\nContingency Table: ", var, " vs y\n")
  print(table(bank[[var]], bank$y))
}

# default
mytab <- table(bank$default, bank$y)
mytab
# mosaic style plots
plot(y~default, data=bank, col=c("black","skyblue"))

ggplot(bank,aes(x=default, fill=y))+geom_bar(position="fill")

# Just to plot what we care about
g <- bank %>%
  group_by(default, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[3:4,], aes(x=default, y=freq, colour=default)) +
  geom_bar(aes(fill=default), stat = "identity") +
  ylab("Proportion of y") 

epitab(mytab, method="oddsratio", oddsratio="wald", pvalue="chi2", rev="columns")


# housing
mytab <- table(bank$housing, bank$y)
mytab
# mosaic style plots
plot(y~housing, data=bank, col=c("black","skyblue"))

ggplot(bank,aes(x=housing, fill=y))+geom_bar(position="fill")

# Just to plot what we care about
g <- bank %>%
  group_by(housing, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[3:4,], aes(x=housing, y=freq, colour=housing)) +
  geom_bar(aes(fill=housing), stat = "identity") +
  ylab("Proportion of y") 

epitab(mytab, method="oddsratio", oddsratio="wald", pvalue="chi2", rev="columns")

#loan
mytab <- table(bank$loan, bank$y)
mytab
# mosaic style plots
plot(y~loan, data=bank, col=c("black","skyblue"))

ggplot(bank,aes(x=loan, fill=y))+geom_bar(position="fill")

# Just to plot what we care about
g <- bank %>%
  group_by(loan, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[3:4,], aes(x=loan, y=freq, colour=loan)) +
  geom_bar(aes(fill=loan), stat = "identity") +
  ylab("Proportion of y") 

epitab(mytab, method="oddsratio", oddsratio="wald", pvalue="chi2", rev="columns")
```

**INTERPRETATION**
For all binary categorical variables, the "no" has a high proportion of yes of y. Which means when they don't have house loan, personal loan or don't have a credit in defualt, the odds of yes to term of deposit is higher than when they have house loan, peronal loan.  

Odd Ratio
- loan ; 2.02 times
- housing: 2.40 times
- default: 1.96 times


## EDA: Categorical vaiables: multi level. 
```{r}
multi.categ.var <- c("job", "marital", "education", "contact", "month", "poutcome")

# Convert each variable in multi.categ.var to factor
bank[multi.categ.var] <- lapply(bank[multi.categ.var], as.factor)

# Optional: verify conversion
str(bank[multi.categ.var])


#job
mytab <- table(bank$job, bank$y)
mytab

# Just to plot what we care about
g <- bank %>%
  group_by(job, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[13:24,], aes(x=job, y=freq, colour=job)) +
  geom_bar(aes(fill=job), stat = "identity") +
  ylab("Proportion of y") 

#marital
mytab <- table(bank$marital, bank$y)
mytab

# Just to plot what we care about
g <- bank %>%
  group_by(marital, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[4:6,], aes(x=marital, y=freq, colour=marital)) +
  geom_bar(aes(fill=marital), stat = "identity") +
  ylab("Proportion of y") 


#education
mytab <- table(bank$education, bank$y)
mytab

# Just to plot what we care about
g <- bank %>%
  group_by(education, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[5:8,], aes(x=education, y=freq, colour=education)) +
  geom_bar(aes(fill=education), stat = "identity") +
  ylab("Proportion of y") 



#contact
mytab <- table(bank$contact, bank$y)
mytab

# Just to plot what we care about
g <- bank %>%
  group_by(contact, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[4:6,], aes(x=contact, y=freq, colour=contact)) +
  geom_bar(aes(fill=contact), stat = "identity") +
  ylab("Proportion of y") 

#month
mytab <- table(bank$month, bank$y)
mytab

# Just to plot what we care about
g <- bank %>%
  group_by(month, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[13:24,], aes(x=month, y=freq, colour=month)) +
  geom_bar(aes(fill=month), stat = "identity") +
  ylab("Proportion of y") 

#poutcome
mytab <- table(bank$poutcome, bank$y)
mytab

# Just to plot what we care about
g <- bank %>%
  group_by(poutcome, y) %>%
  summarise(cnt = n()) %>%
  mutate(freq = round(cnt / sum(cnt), 4)) %>%
  arrange(desc(freq))

g

ggplot(g[5:8,], aes(x=poutcome, y=freq, colour=poutcome)) +
  geom_bar(aes(fill=poutcome), stat = "identity") +
  ylab("Proportion of y") 
```

## correlation Matrix
```{r}
library(GGally)

# ggpairs with color
ggpairs(
  bank,
  columns = c("y.num", "age", "balance", "day", "duration", "campaign", "pdays", "previous"),
  mapping = aes(color = y)
)
```

## Boundaries classification(Interaction or polynomial or additive)
```{r}
```


# Objective 1: build a logistic regression model for interpretaion purposes using the training data


```{r}
# base
baselogistic <- glm(y~age+balance+day+duration+campaign+pdays+previous+job+marital+education+contact+month+poutcome+housing+loan+default, data=train, family = "binomial")
summary(baselogistic)


```
**Interpretation**
Age, pdays, previous and default has large p-values. AIC is 2648. Take those out.

```{r}
# Take out large p-value
logistic1 <- glm(y~balance+day+duration+campaign+job+marital+education+contact+month+poutcome+housing+loan, data=train, family = "binomial")
summary(logistic1)

# Adding logs
logistic2 <- glm(y~balance+day+duration+campaign+job+marital+education+contact+month+poutcome+housing+loan, data=train, family = "binomial")
summary(logistic2)

```


# Do we have to do something with unknown data?

## Interaction terms
```{r}

# balance
ggplot(bank,aes(x=balance, y=y.num, colour=housing)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~housing)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=marital)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~marital)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=job)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~job)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=education)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~education)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=loan)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~loan)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=default)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~default)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=contact)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~contact)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=month)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~month)+ ylim(-.2,1.2)

ggplot(bank,aes(x=balance, y=y.num, colour=poutcome)) +geom_point() + geom_smooth(method="loess", size=1, span=1.5) +facet_wrap(~poutcome)+ ylim(-.2,1.2)
```